---
title: "FluSight Hubverse Cloud Vignette"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Library and system set up
If you don't have the following packages, make sure to install them

```{r install, eval=FALSE}
remotes::install_github("Infectious-Disease-Modeling-Hubs/hubData")
remotes::install_github("Infectious-Disease-Modeling-Hubs/hubEnsembles")
remotes::install_github("Infectious-Disease-Modeling-Hubs/hubVis")
```

Once the required packages are installed, load the following:

```{r libraries, message=FALSE}
library(hubData)
library(hubEnsembles)
library(hubVis)

library(arrow)
library(dplyr)
```


## Load the data
CDC's FluSight Forecast Hub, which began as a [GitHub repository](https://github.com/cdcepi/FluSight-forecast-hub/tree/main), has been mirrored to the cloud via a publicly-accessible AWS S3 bucket. This means that instead of relying on a local clone, the data can be directly accessed from the cloud.

In particular, `hubData` makes it possible to access the model output data as an already-formatted tibble in a few simple steps.

```{r}
hub_path_cloud <- s3_bucket("cdcepi-flusight-forecast-hub/") # connect to bucket
hub_con <- connect_hub(hub_path_cloud, file_format = "parquet") # connect to hub
data_cloud <- hub_con |>
  collect() # collect all model output into single tibble
```

This collected model output can then be changed with the common `dplyr` operations as desired

```{r}
# perform various operations on data
filtered_outputs <- data_cloud |>
  filter(output_type == "quantile", location == "US", horizon > -1) |>
  mutate(output_type_id = as.numeric(output_type_id)) |>
  as_model_out_tbl()

# print data
head(filtered_outputs)
```

Currently, there is no `hubData` integration to extract target data from the cloud (though this will likely be supported in the future), so instead we load it directly from a local file. This file contains observed influenza hospital admissions up through the final reference date of the 2023-2024 influenza season. This file is stored in the "vignettes/articles" directory of the `hubData` package.

```{r, message=FALSE}
target_data <- readr::read_csv("target-hospital-admissions_2024-04-27.csv") |>
  select(date, location, value) # keep only required columns

head(target_data) # print
```

## Calculate some ensembles
See [hubEnsembles package](https://infectious-disease-modeling-hubs.github.io/hubEnsembles/index.html) for more information

Quantile mean: 
```{r}
mean_ens <- filtered_outputs |>
  hubEnsembles::simple_ensemble(model_id = "mean-ensemble")
head(mean_ens)
```

Linear pool (with normal tails): 
```{r}
linear_pool <- filtered_outputs |>
  hubEnsembles::linear_pool(model_id = "linear-pool")
head(linear_pool)
```

## Plot the output
See [hubVis package](https://infectious-disease-modeling-hubs.github.io/hubVis/index.html) for more information

### Data processing
Modify and filter the target data for plotting:

```{r}
target_data_plotted <- target_data |>
  mutate(target = "wk inc flu hosp", observation = value) |>
  filter(location == "US", date >= as.Date("2023-09-23"))
```

Modify and filter forecast data for plotting
```{r}
reference_dates <- unique(filtered_outputs$reference_date)
model_outputs_plotted <- filtered_outputs |>
  filter(model_id %in% c("FluSight-baseline", "MOBS-GLEAM_FLUH", "PSI-PROF")) |>
  rbind(mean_ens, linear_pool) # bind with ensembles
```

### Plot
Single plot, single set of forecasts for one reference date

```{r}
model_outputs_plotted |>
  filter(reference_date == as.Date("2024-04-20")) |>
  hubVis::plot_step_ahead_model_output(
    target_data_plotted,
    x_col_name = "target_end_date",
    use_median_as_point = TRUE,
    group = "reference_date",
    interactive = FALSE
  )
```

Single plot, multiple sets of forecasts for reference dates every 4 weeks
```{r}  
model_outputs_plotted |>
  filter(reference_date %in% reference_dates[seq(3, 31, 4)]) |>
  hubVis::plot_step_ahead_model_output(
    target_data_plotted,
    x_col_name = "target_end_date",
    use_median_as_point = TRUE,
    group = "reference_date",
    interactive = FALSE
  )
```

Faceted plot, multiple sets of forecasts for reference states every 4 weeks

```{r}
model_outputs_plotted |>
  filter(reference_date %in% reference_dates[seq(3, 31, 4)]) |>
  hubVis::plot_step_ahead_model_output(
    target_data_plotted,
    x_col_name = "target_end_date",
    use_median_as_point = TRUE,
    show_legend = FALSE,
    facet = "model_id",
    group = "reference_date",
    interactive = FALSE
  )
```
